Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis

### 2023-04-03 Write-ups/Manuscript.tex
spell checks

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{April 01, 2023}

@@

Biomedical investigators are often interested in predicting future
observations of subjects based on their historical data, typically
referred to as dynamic prediction. Traditionally, this type of data has
been modeled using marginal models (generalized estimating equations) or
conditional models (mixed effect models)
(\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
\protect\hyperlink{ref-liang1986}{LIANG and ZEGER 1986};
\protect\hyperlink{ref-lindstrom1990}{Lindstrom and Bates 1990};
\protect\hyperlink{ref-davidian2003}{{``Nonlinear models for repeated
measurement data''} 2003}), and predictions are made based on the
correlation between repeated measures from the same subject, and/or
covariates that can be either fixed or time-varying. However, these
methods are limited in terms of flexibility of correlation structure and
the ability to handle out-of-sample prediction. When sample size is
large or the density of repeated measures is high, they also tend to
cause severe computational burden
(\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}).

To address these problems, one may turn to functional mixed effect
models instead when measures are dense across the domain of measures.
Such methods accommodate more flexible correlation structure by modeling
subject-specific random effects as a function, and non-parametric
smoothing (\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}) can
be incorporated to speed up the computation, such as spline basis
functions or eigenfunctions from functional principal component analysis
(FPCA). The introduction of basis functions also makes out-of-sample
prediction more straightforward. Instead of estimating subject-specific
random effects of new observations, we can simply estimate
coefficients/loadings on the basis function used for smoothing. Existing
research on dynamic prediction with functional data analysis methods has
been focusing on continuous/Gaussian outcomes, modelling
subject-specific random effects with FPCA
(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
predict missing observations in partially observed function tracks, and
Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
similar goals using Markov Chains. While methods mentioned above used
only partial observations for prediction with an intercept-only model,
Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018}) proposed
Functional Concurrent Regression (FCR) framework which can incorporate
the effect of subject-specific predictors.

Unfortunately, fewer papers have focused on its extension to
non-Gaussian data, such as series of binary or count outcomes. Existing
methods also tend to be very computationally intensive. For example,
Chen et al. (\protect\hyperlink{ref-chen2013}{2013}) proposed approaches
to fit marginal functional models that is compatible to multi-level,
generalized outcomes. Goldsmith et al.
(\protect\hyperlink{ref-goldsmith2015}{2015}) established a model
framework that takes into account the fixed effect of time-invariant
covariates, with parameters estimated with Bayesian method in
\emph{Stan}. Gertheiss et al.
(\protect\hyperlink{ref-gertheiss2016}{2016}) identified bias introduced
by directly applying FPCA methods to generalized functions, and proposed
to address this problem using a two-stage, joint estimation strategy.
Linde (\protect\hyperlink{ref-linde2019}{2009}) used an adapted Bayesian
variational algorithm for FPCA of binary and count data. In terms of
implementation, Wrobel et al. (\protect\hyperlink{ref-wrobel2019}{2019})
proposed a fast, efficient way to fit GFPCA on binary data using EM
algorithm, accompanied by the an open source R package \emph{registr}.

In this paper, we aim to develop a fast, scalable method for dynamic
prediction of discrete function tracks based on functional mixed effect
model with FPCA smoothing. Section 2 presents the procedure of the
proposed method. In Section 3, we illustrate the performance and
efficiency of our proposed method in a simulation study. In Section 4,
we apply this method to a real-world dataset. Section 5 presents a
discussion of advantages and limitation of the proposed method.

@@

The observed data for a single subject \(i\) is (\(t\), \(Y_i(t)\)),
where \(t\) consists of dense, discrete points along the functional
domain, and \(Y_i(t)\) is the non-Gaussian outcome observed at \(t\). We
assume that the outcome \(Y_i(t)\) can be characterized by a latent
continuous function \(\eta_i(t)\). That is, at a specific t, \(Y_i(t)\)
follows a exponential family distribution such that:

@@

According to the Karhunen-Loève theorem, the unobserved random effect
function \(b_i(t)\) can be represented as the linear combination of a
infinite set of orthogonal functions:
\(b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)\). We approximate this
function by a finite set of \(\phi_{k}(t)\), such that
\(b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)\), where
\(\epsilon_i(t)\) indicates random noise introduced by this
approximation, following a zero-mean Gaussian process. At each specific
point t, \(\epsilon_i(t) \sim N(0, \sigma^2)\).

In practice, this approximation could be implemented through an FPCA
procedure on \(\eta_i(t)\). In this case
\(\phi_{k}(t),\ k \in \{1,...,K\}\) are the first K eigenfunctions that
explain the most variation in \(\eta_i(t)\), and \(\xi_{ik}\) are
subject-specific PC scores (or loadings) on each eigenfunction.
Additionally, \(\xi_{ik}\) are mutually independent over both subject
(\(i\)) and eigenfunctions (\(k\)). That is, each \(\xi_{ik}\) follows
normal distribution \(N(0, \lambda_k)\) where \(\lambda_k\) is the kth
eigenvalues.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Bin the observed outcomes in to small, non-overlapping, equal length
  intervals. We hereafter index the bins by their midpoints \(s\).
\item
  Fit a local Generalized Mixed Model at every bin. Specifically, at bin
  \(s\), we fit \(g[E(Y_i(t))] = \beta_0(s)+b_i(s)\) for all \(t\) in
  bin \(s\). From this series of models we can estimate the individual
  latent functions at every bin:
  \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
\item
  Fit FPCA on the estimated latent functions:
  \(\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)\),
  and obtain estimates of basis functions
  \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
  \(\hat{\lambda}_1...\hat{\lambda}_k\), mean function \(\hat{f}_0(s)\)
  and residual variance \(\hat{\sigma}^2\).
\end{enumerate}

@@

In this section, we illustrate the predictive performance and
computational efficiency of the proposed method through a simulation
study. We simulated 50 datasets, each with 500 subjects. For every
subject, we generate 1000 binary outcomes \(Y_i(t) \in \{0, 1\}\) across
functional domain \(t \in [0, 1000]\), where the distribution of outcome
is characterized by a continuous latent function. The data generation
mechanism can be expressed as follows:

@@

In this simulation, we set \(f_0(t) = 0\). \(\xi_{ik}\) are mutually
independent normal random variables \(\xi_{ik}\sim N(0, \gamma_k)\).
Here we set the values of \(\gamma_k\) to be \(0.5^{k-1}\),
\(k \in (1,..., 4)\). In addition, for simplicity of presentation, we
generate data on a regular grid, which means observations points are
equally distributed across \([0, 1000]\) and are the same for all
subjects.

We use two metrics to evaluate the out-of-sample predictive performance:
integrated squared error (ISE) and
Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction
accuracy of latent continuous function. It is evaluated on the binned
grid at midpoints of each unobserved bin. If the entire functional
domain has S bins, but we have observations up to the mth bin, then ISE
is defined as
\(\frac{1}{N}\sum_{i=1}^N\sum_{s=s_m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2\).
The second metric, AUC, focuses on evaluation of prediction of the
binary outcome. Since the binary outcomes are generated on the original,
un-binned grid, we evaluated AUC on this grid and estimated values of
latent functions between bins with linear interpolation.

@@

The average ISE and AUC across all simulation is presented in Table 1,
and Figure 1 presents the predicted function curves of four
randomly-drawn subjects from the first simulation. The prediction is
conditioning on different length of observed track (specifically, with
observations up to t = 200, 400, 600, 800 respectively), and evaluation
is made on equal-length time windows on the unobserved tracks following
the maximum observation time. As Table 1 reveals, our proposed method
(fGFPCA) outperformed GLMMadaptive under every scenario, which was
expected since GLMMadaptive can only accommodate simple model
structures. In addition, fGFPCA also took less computation time. For one
simulated dataset, fGFPCA spent 4.94 minutes on model fitting and
out-of-sample prediction, while GLMMadaptive took 5.11 minutes.

Because of the flexibility of our proposed model framework, the accuracy
of prediction at specific time points would improve with more observed
data. This is revealed in Table 1 as ISE decreases and AUC increases
with maximum observed time (left-to-right), also in Figure 1 as
predicted curves get closer with longer observed track. However the same
tendency is not observed with GLMMadaptive models, as a result of
restricted model flexibility. A linear model that fits well to a
specific part of latent function can fit very badly to the following
parts, especially when the underlying latent function has cyclic
patterns. Therefore, more observations do not necessarily make the model
more predictive. Since the latent Gaussian function tracks are not known
in this case, prediction performance on evaluated with AUC along.

Table 2 compares the predictive performance of both models. When the
observed time is short, the two method performend similarly when
predicting recent days, however fGFPCA did better in predicting further
days along the trial. As more observations coming in, fGFPCA sees
increasingly higher AUC compared to GLMMadaptive model. This could be a
result of the difference of model flexibility of the two methods. While
the linear relationship in underlying latent functions are not likely to
change immediately following the observed track, the whole track may
need to be much more flexible, which is a better fit for fGFPCA model.
This possible explanation is illustrated in Figure 2 using predicted
latent function tracks of four randomly drawn subjects.

@@

We will illustrate the performance of our proposed method using data
from a randomized trial on weight loss at University of Colorado
Anschutz Medical Campus (\protect\hyperlink{ref-bothwell2022}{Bothwell
et al. 2022}; \protect\hyperlink{ref-ostendorf2022}{Ostendorf et al.
2022}). The study originally aimed to study the effect of dietary
strategies on weight loss, and participants were given a smart scale for
daily at-home weighing. This record was then scaled into a binary
indicator of participant adherence to the daily-weigh-in procedure,
where 1 indicates participants followed the requirement and 0 otherwise.
In this section, we will use this series of daily indicators as the
binary functional outcome, and use the proposed method to dynamically
predict participant adherence on future days given historical track.

The dataset includes 55 participants aged 22-56 at baseline. Adherence
to daily weigh-in is availabe four 400 consecutive days for all
participants. In other words, for each individual, 400 binary
measuresments were taken on a regular grid along the functional domain
(number of days into the study). We will fit both fGFPCA and
GLMMadaptive models and compared their performance on this data. For
fGFPCA, every 10 observations are binned together, resulting in 40 bins
for final model fitting. Linear interpolation was used for extension
back to the unbinned grid.

@@

The simulation study and case study of daily weigh-in data above have
revealed the proposed method can achieve better predictive performance
with much less computational burden. Compared to existing method based
on linear mixed model, fGFPCA can accomodate flexible correlation
struture between repeated measure, as well as high density of
measurement. This is particularly useful when underlying outcome shows
cyclic pattern, such as the simulation setting in Section 3. While the
binary outcomes are generated from a cyclic Gaussian process, fGFPCA has
much lower ISE and higher AUC than GLMM. Its prediction accuracy also
improves as more outcomes are observed, which is a desirable property
for dynamic prediction where the datasets would be periodically updated
with newly collected data.

>>>>

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{April 03, 2023}

@@

Biomedical investigators are often interested in predicting future
observations of subjects based on their historical data, typically
referred to as dynamic prediction. Traditionally, this type of data has
been modeled either marginally, such as generalized estimating
equations, or constitutionally on specific subjects, such as mixed
effect models (\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
\protect\hyperlink{ref-liang1986}{LIANG and ZEGER 1986};
\protect\hyperlink{ref-lindstrom1990}{Lindstrom and Bates 1990};
\protect\hyperlink{ref-davidian2003}{{``Nonlinear models for repeated
measurement data''} 2003}). For subject-specific models, predictions are
made based on the correlation between repeated measures from the same
subject and covariates that can be either fixed or varying. However,
most of these methods are based on linear model structure, which is very
limited in terms of flexibility of correlation. When sample size is
large or the density of repeated measures is high, they also tend to
cause severe computational burden
(\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}). In addition,
with generalized, non-Gaussian outcomes, these methods lack the ability
to handle out-of-sample prediction because there is usually no
close-form solution for subject-specific random effects

To address these problems, one may turn to functional mixed effect
models instead when measures are dense across the domain of measures.
Such methods accommodate more flexible correlation structure by modeling
subject-specific random effects as a function. Non-parametric smoothing
(\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}) are often
incorporated to speed up the computation, where random effects are
approximated by spline basis functions or eigenfunctions from functional
principal component analysis (FPCA). The introduction of basis functions
also makes out-of-sample prediction more straightforward. Instead of
estimating subject-specific random effects of new observations, we can
simply estimate coefficients/loadings on the basis function used for
smoothing. Existing research on dynamic prediction with functional data
analysis methods has been focusing on continuous/Gaussian outcomes,
usually modelling subject-specific random effects with FPCA
(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
predict missing observations in partially observed function tracks, and
Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
similar goals using Markov Chains. While methods mentioned above used
only partial observations for prediction with an intercept-only model,
Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018}) proposed
Functional Concurrent Regression (FCR) framework which can incorporate
the effect of subject-specific predictors.

Unfortunately, fewer papers have focused on its extension to
generalized, non-Gaussian outcomes, such as series of binary or count
outcomes. Existing methods also tend to be very computationally
intensive. For example, Chen et al.
(\protect\hyperlink{ref-chen2013}{2013}) proposed approaches to fit
marginal functional models that is compatible to multi-level,
generalized outcomes. Goldsmith et al.
(\protect\hyperlink{ref-goldsmith2015}{2015}) established a model
framework that takes into account the fixed effect of time-invariant
covariates, with parameters estimated with Bayesian method in
\emph{Stan}. Gertheiss et al.
(\protect\hyperlink{ref-gertheiss2016}{2016}) identified bias introduced
by directly applying FPCA methods to generalized functions, and proposed
to address this problem using a two-stage joint estimation strategy.
Linde (\protect\hyperlink{ref-linde2019}{2009}) used an adapted Bayesian
variational algorithm for FPCA of binary and count data. In terms of
implementation, Wrobel et al. (\protect\hyperlink{ref-wrobel2019}{2019})
proposed a fast, efficient way to fit GFPCA on binary data using EM
algorithm, accompanied by the an open source R package \emph{registr}.

In this paper, we aim to develop a fast, scalable method for dynamic
prediction of generalized functional outcome based on functional mixed
effect model with FPCA smoothing. Specifically, ``Generalized''
indicates that the outcome follows non-Gaussian exponential family
distribution, and the term ``functional'' indicates that repeated
measure are collected densely across the study domain (e.g., time).
Section 2 explains the implementation procedure of the proposed method.
In Section 3, we illustrate the performance and efficiency of our
proposed method in a simulation study. In Section 4, we apply this
method to a real world dataset from a weight loss study. Section 5
presents a discussion of advantages and limitation of the proposed
method.

@@

The observed data for a single subject \(i\) is (\(t\), \(Y_i(t)\)),
where \(t\) consists of dense, discrete points along the domain of data
collection. In practice, \(t\) is usually a time index along the
duration of study. We hereafter refer to the range of \(t\) as the
``functional domain''. For simplicity, we further assume that outcomes
are collected on a regular grid over the functional domain, meaning that
\(t\) are evenly distributed and does not change across subjects.

\(Y_i(t)\) is the non-Gaussian outcome observed at \(t\). We assume that
the outcome \(Y_i(t)\) can be characterized by a latent continuous
function \(\eta_i(t)\). That is, at a specific t, \(Y_i(t)\) follows a
exponential family distribution such that:

@@

According to the Karhunen-Loève theorem, the unobserved random effect
function \(b_i(t)\) can be represented as the linear combination of a
infinite set of orthogonal functions:
\(b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)\). We approximate this
function by a finite set of \(\phi_{k}(t)\), such that
\(b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)\), where
\(\epsilon_i(t)\) indicates random noise introduced by this
approximation. This random noise is assumed to follow a zero-mean
Gaussian process, such that at each specific point t,
\(\epsilon_i(t) \sim N(0, \sigma^2)\).

In practice, this approximation could be implemented through an FPCA
procedure on \(\eta_i(t)\). In this case
\(\phi_{k}(t),\ k \in \{1,...,K\}\) are the first K eigenfunctions that
explain the most variation in \(\eta_i(t)\). These eigenfunctions, also
referred to as Principal Component (PC) functions, are orthogonal are
can be orthonormal with proper scaling. \(\xi_{ik}\) are
subject-specific scores (or loadings) on each PC function. According to
properties of FPCA method, these scores are mutually independent over
both subject (\(i\)) and eigenfunctions (\(k\)). That is, each
\(\xi_{ik}\) follows normal distribution \(N(0, \lambda_k)\), where
\(\lambda_k\) is the kth eigenvalues.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Bin the observed outcomes in to small, non-overlapping, equal length
  intervals. We hereafter index the bins by their midpoints \(s\). The
  number of observations from the same subject in bin \(s\) is \(n_s\),
  and \(t_{sj}\) refers to the jth observation point in bin \(s\).
\item
  Fit a local Generalized Mixed Model at every bin. Specifically, at bin
  \(s\), we fit \(g[E(Y_i(t_{sj}))] = \beta_0(s)+b_i(s)\),
  \(j\in\{1, ..., n_s\}\). From this series of models we can estimate
  the individual latent functions at every bin:
  \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
\item
  Fit FPCA on the estimated latent functions:
  \(\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)\),
  and obtain estimates of basis functions
  \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
  \(\hat{\lambda}_1...\hat{\lambda}_k\), mean function \(\hat{f}_0(s)\)
  and residual variance \(\hat{\sigma}^2\).
\end{enumerate}

@@

In this section, we illustrate the predictive performance and
computational efficiency of the proposed method through a simulation
study. We simulated 50 datasets, each with 500 subjects. For every
subject, we generate 1000 binary outcomes \(Y_i(t) \in \{0, 1\}\) across
functional domain \(t \in (0, 1000]\), where the distribution of outcome
is characterized by a continuous latent function. The data generation
mechanism can be expressed as follows:

@@

In this simulation, we set \(f_0(t) = 0\). \(\xi_{ik}\) are mutually
independent normal random variables \(\xi_{ik}\sim N(0, \lambda_k)\).
Here we set the values of \(\lambda_k\) to be \(0.5^{k-1}\),
\(k \in (1,..., 4)\). In addition, for simplicity of presentation, we
generate data on a regular grid, which means observations points are
equally distributed across \((0, 1000]\) and are the same for all
subjects.

We use two metrics to evaluate the out-of-sample predictive performance:
integrated squared error (ISE) and
Area-Under-the-Receiver-Operator-Curve (AUC). ISE assesses the
prediction accuracy of latent continuous functions. It is evaluated on
the binned grid at midpoints of each unobserved bin. With observations
up to bin \(s_m\), ISE is defined as
\(\frac{1}{N}\sum_{i=1}^N\sum_{s>s_m} (\hat{\eta}_i(s)-\eta_i(s))^2\).
The second metric, AUC, focuses on evaluation of prediction of the
binary outcome. Since the binary outcomes are generated on the original,
un-binned grid, we evaluated AUC on this grid by estimating values of
latent functions between bins with linear interpolation.

@@

The average ISE and AUC across all simulation is presented in Table 1,
and Figure 1 presents the predicted latent function of four randomly
selected subjects from the first simulated dataset. The prediction is
made conditioning on different length of observed track (specifically,
with observations up to t = 200, 400, 600, 800 respectively), and
evaluation is made on equal-length time windows on the unobserved tracks
following the maximum observation time. As Table 1 reveals, the proposed
method (fGFPCA) outperforms GLMMadaptive under every scenario, which was
expected since the linear model structure of GLMM is too simple for the
cyclic data generation mechanism. In addition, fGFPCA also took less
computation time. For one simulated dataset, fGFPCA spent 4.94 minutes
on model fitting and out-of-sample prediction overall, while
GLMMadaptive took 5.11 minutes.

Because of the flexibility of our proposed model framework, the accuracy
of prediction at specific time points would improve as more observed
data is collected. This is revealed in Table 1 as ISE decreases and AUC
increases with maximum observed time (left-to-right), also in Figure 1
as predicted curves get closer to the true latent function with longer
observed track. However the same tendency is not observed with
GLMMadaptive models, as a result of restricted model flexibility. A
linear model that fits well to a specific part can fit very badly to
other parts along the underlying latent function tracks, especially when
the shapes of functions are wiggly or cyclic. Therefore, more
observations do not necessarily make the model more predictive.

@@

In this section, we illustrate the performance of our proposed method
using data from a randomized trial on weight loss at University of
Colorado Anschutz Medical Campus
(\protect\hyperlink{ref-bothwell2022}{Bothwell et al. 2022};
\protect\hyperlink{ref-ostendorf2022}{Ostendorf et al. 2022}). The study
originally aimed to study the effect of dietary strategies on weight
loss, and participants were given a smart scale for daily at-home
weighing. This record was then scaled into a binary indicator of
participant adherence to the daily-weigh-in procedure, where 1 indicates
participants followed the requirement and 0 otherwise. In this section,
we will use this series of daily indicators as the binary functional
outcome, and use the proposed method to dynamically predict participant
adherence on future days given historical track.

The dataset includes 55 participants aged 22-56 at baseline. Adherence
to daily weigh-in is available along 400 consecutive days for all
participants. In other words, for each individual, 400 binary
measurements were taken on a regular grid along the functional domain
(number of days into the study). We will fit both fGFPCA and
GLMMadaptive models and compared their performance on this data. For
fGFPCA, every 10 observations are binned together, resulting in 40 bins
for final model fitting. Linear interpolation was used for extension
from the binned grid back to the un-binned grid. Since the latent
continuous function tracks are not known in this case, prediction
performance on evaluated with AUC alone.

Table 2 compares the predictive performance of both models. Similar to
the simulatio study, prediction is made conditioning on observations up
to the 100th, 200th and 300th day respectively, and evaluation is made
every 100 days following the maximum observation time. As is revealed,
when the observed time is short (up to 100th day), the two method
performed similarly when predicting recent days, however fGFPCA did
better in predicting further days along the trial (prediction time
window of \((300, 400]\)). As more observations coming in, fGFPCA sees
increasingly higher AUC compared to GLMMadaptive model. This could be a
result of the difference in model flexibility of the two methods. While
the linear relationship in underlying latent functions are not likely to
change immediately following the observed track, the whole track could
be much more complex. Therefore, while fGFPCA performs similarly to
GLMMadaptive right after the end of observation, it does a much better
job capturing the underlying pattern along the whole track. Figure 2
further illustrate such difference using predicted latent function
tracks of four randomly drawn subjects.

@@

The simulation study and case study of daily weigh-in data above have
revealed the proposed method can achieve better predictive performance
with much less computational burden. Compared to existing methods based
on linear mixed model, fGFPCA can accommodate flexible correlation
structure between repeated measure, which is particularly useful when
outcome is generated from latent functions with wiggly or cyclic
pattern. As in the simulation setting in Section 3, the binary outcomes
are generated from a cyclic process. fGFPCA has much lower ISE and
higher AUC than GLMM in this case, and its prediction accuracy also
improves as more outcomes are collected, which is a desirable property
for dynamic prediction where the datasets would be periodically updated
with newly collected data. Moreover, fGFPCA also significantly reduce
time spent on model fitting and predict, which is beneficial when
repeated measures are dense or sample size is large. In fact, as far as
we are aware of, the proposed method is the only feasible method that
can handle the scale of the simulated datasets with compatible
flexibility.
